---
title: "NHL Analysis"
author: "Ben Smith and Garrett Wallace"
date: "April 4, 2018"
output: html_document
---
```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

#NHL Player Salary Analysis and Prediction
###Ben Smith and Garrett Wallace





##Step 1: Looking at the Data (and a little cleaning)

```{r}
nhl <- read.csv("nhl.csv")
str(nhl)
```

```{r echo=FALSE}
nhl$Position <- gsub("/.*","",nhl$Position)
levels(nhl$Position) <- c(levels(nhl$Position), "W")
nhl$Position[nhl$Position == "RW"] <- "W"
nhl$Position[nhl$Position == "LW"] <- "W"
nhl$Position <- as.factor(nhl$Position)

nhl$DftRd[is.na(nhl$DftRd)] <- mean(nhl$DftRd, na.rm=TRUE)
nhl$City <- NULL
nhl$Pr.St <- NULL
nhl$Cntry <- NULL

library(lubridate)
nhl$Born <- ymd(as.character(nhl$Born))
nhl$Born[is.na(nhl$Born)] <- mean(nhl$Born, na.rm=TRUE)

nhl$Last.Name <- NULL
nhl$First.Name <- NULL

NHL$BirthYear <- as.factor(as.POSIXlt(NHL$Born, format="%Y-%m-%d")$year)
NHL$BirthYear <- as.numeric(levels(NHL$BirthYear))[NHL$BirthYear]
NHL$BirthYear <- (NHL$BirthYear + 3800)

  
#is.na(NHL$DftYr) <- NHL$BirthYear + 20
NHL$DftYr[is.na(NHL$DftYr)] <- 0
for (x in NHL$DftYr) {
  if (x == 0) {
    x = NHL$BirthYear + 20
  }
}

nhlc <- subset(nhl, nhl$Position=="C")
nhlw <- subset(nhl, nhl$Position=="W")
nhld <- subset(nhl, nhl$Position=="D")
```


Looking initially at our data, we see plenty of different opportunities to analyze it. There are over 150 different variables on which to analyze hockey players, and our objective will be to see how these relate to the overall quality of the player and thus the amount they are paid.

Neither Garrett nor Ben are particularly adept at hockey knowledge, so we turned to some friends and the Internet for advice on how to best begin parsing through all the variables. After some discussion, we broke player positions down to three categories that tend to have the same traits; Defensemen, Centers, and Wingsmen. Our ultimate intention is to build regressions both on players as a whole and within those subsets to extract meaningful data. We also got rid birth city/state/country as a predictor of salary, but kept nationality; it is possible people need to be paid more or less to make it a competitive wage relative to their home country. 

When selecting which variables to narrow down for subsets and for the entire population, we abided by general principles:
- Goal scoring would be generally attributed to offensive players (centers and wings) and goals allowed to defensive players. Even though there is overlap between the positions, the majority of goals are scored by offensive players and majority of goals are allowed by defenders.
- Disregarded specific “when trailing", "when winning", and "when tied” statistics as well as detailed descriptions of shots. For a general model this would overcomplicate things, but in the future if a further analysis wanted to be done on these specific situations, our general code could be adapted to do so. For example, if a team wanted to examine a player's percentage of faceoffs won when trailing vs when winning, our code would allow them to do so.
- Avoid confounding variables as much as possible. For example, shots on goal offensively is loosely related to shots against defensively - when your team is shooting, you are inherently not allowing any shots.

Just for the sake of curiosity, lets get a few basic metrics:


```{r echo=FALSE}

table(nhl$Position)

cat("NHL Average:", mean(nhl$Salary), "", sep ="\n")

cat("Center Average:", mean(nhlc$Salary), "", sep ="\n")

cat("Wingman Average:", mean(nhlw$Salary), "", sep ="\n")

cat("Defense Average:",mean(nhld$Salary), "", sep ="\n")
```



So it looks like there's a little variability among players in salary. Center's are closest to average, with defensive players making a bit more and wings making a bit less. We also see that wings are the most readily available number of players in the dataset, which (assuming this dataset is representative of the NHL) might mean there are more wingsmen overall and thus any one wingsman is on average less valuable. 

Now that we have some rough subsets, lets do some high-level analysis. 



##Initial Analysis

```{r}
allvars <- reformulate(setdiff(colnames(nhl), "Salary"), response = "Salary")

lin1 <- lm(allvars, data=nhl)
summary(lin1)
```



####That's a lot of variables!

We can see that when looking at the groups in aggregate, the data that's actually significant to the model is pretty sparse relative to all other inputs. Still, though, the Adjusted R-Squared of .6814 suggests that this data does a vaguely acceptable job of predicting.

Still, we're unsure that linear regression is the best way to look at it (and when using other models, 150 variables for an 874-row dataset might not produce the most reliable data. So, from here we'll implement the advice we got on which statistics matter more, and we'll also try a few other models.


  
###2a: Offensive Players Analysis

While NHL fans praise hockey players for their defensive toughness, you need good offense to win games. As mentioned earlier, centers and wings account for the majority of attacking plays that can end up in goals, and we want to identify exactly which factors NHL teams most value in their offensive players, thus resulting in the greatest salaries.

The analysis of all offensive players combines stats that are important for both centers and wings. Centers are generally responsible for faceoffs, so the analysis of wings does not include faceoff statistics but is the same otherwise.

```{r, echo=FALSE}

offense_analysis <- nhlc[,c("Salary", "X.FOT", "dzFOL", "dzFOW", "nzFOL", "nzFOW", "ozFOL", "ozFOW", "A1", "A2", "CF", "DSF", "FF", "G", "GF", "iCF", "iDS", "iFF", "iFOL", "iFOW", "iSCF", "ixG", "Pass", "RSF", "SCF", "SH.", "FO.")]

#nzGFPF OCF OFF OppCF60 OppFF60 OppGF60 OppSF60 TMGF60

offensevars <- reformulate(setdiff(colnames(offense_analysis), "Salary"), response = "Salary")

offensemodel <- lm(offensevars, data=nhl)
summary(offensemodel)
```

```{r, echo=FALSE}

wings_analysis <- nhlc[,c("Salary", "A1", "A2", "CF", "DSF", "FF", "G", "GF", "iCF", "iDS", "iFF", "iSCF", "ixG", "Pass", "RSF", "SCF", "SH.")]

wingsvars <- reformulate(setdiff(colnames(wings_analysis), "Salary"), response = "Salary")

wingsmodel <- lm(wingsvars, data=nhl)
summary(wingsmodel)
```

In the linear models for all offensive players and exclusively wings, the only variable that came back as statistically signficiant with a p-value of .001 was iDS - dangerous shots taken by this player, the sum of rebounds and shots off the rush. When considered logically, this makes sense when determining what teams are paying for in regards to offensive players - they want individuals who are going to get the puck on net with a high quality shot and not just aimlessly shoot and hope for the best, in addition to being aggressive off rebounds. When players have a high degree of danger shots, it minimizes the importance of total quantity of shots.

For wings, CF (the team's shot attempts when this player was on the ice) and primary assists (passes to a player that scored) were also significant. Overall, we can conclude that teams are paying all offensive players for danger shots, but also expect wings to contribute to their team's offensive performance in ways other than shooting. We were surprised to see that faceoff statistics were not considered to be statistically significant.

``` {r, echo=FALSE}
normalize <- function(x) { 
  return((x - min(x)) / (max(x) - min(x)))
}

offense_MM <- as.data.frame(model.matrix(~ . -1, data = offense_analysis))
offense_MM$Salary <- NULL

offense_norm <- as.data.frame(lapply(offense_MM, normalize))
#head(offense_norm)

nrow(offense_norm)
nrow(offense_analysis)

offense_norm$salary <- offense_analysis$Salary
levels(offense_norm$salary) <- c("0", "1")
offense_norm$salary <- as.numeric(offense_norm$salary)
offense_norm$salary <- offense_norm$salary-1
#why do we need previous row

offense_train <- offense_norm[1:194, ]
offense_test <- offense_norm[195:258, ]
#line before may be 251 and not 258 - check this

library(neuralnet)

f <-reformulate(setdiff(colnames(offense_train), "salary") , response = "salary")

offense_ann <- neuralnet(formula = f, data = offense_train, hidden=2, stepmax = 1e6)

plot(offense_ann)

```

``` {r, echo=FALSE}

correlation <- cor(offense_analysis, use = "complete.obs")
round(correlation, 2)
install.packages("Hmisc")
library(Hmisc)
res2 <- rcorr(as.matrix(offense_analysis))
res2
res2$P
```



### Conclusion

Overall, we've identified which factors tend to be the most significant for each position in regard to determining their salary. Now, NHL front offices must determine whether the stats they are paying for are actually worthwhile - if they find that ...


